Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-31 text/paper1/paper.tex
SIM reference, minor tweaks

There is a long tradition for using RR Lyrae stars in the Galaxy to study structure
\citep[going back to first estimates of the distance to the Galactic center][]{shapley18},
substructure \citep[e.g.][]{sesar10} and distances to satellite galaxies [e.g.][]{clementini03}.
However, studies of RR Lyrae at optical wavelengths are limited by both metallicity affects on the intrinsic 
brightness of these stars and variable extinction along the line of sight.
Moreover, systematic differences between instruments make it difficult to tie observations across the sky
to a common scale. 
Hence, while among the best distance indicators available to us today, the absolute uncertainty in
individual estimates are typically no better than 10\%

At longer wavelengths, RR Lyrae promise tighter constraints on distances.
The near-infrared period-luminosity (PL) relation for RR
Lyrae stars was first mapped by \citet{longmore86} 
using $K$-band observations of stars in Galactic
globular clusters and later shown to be due to
the
changing bolometric correction for these stars as a function of the
observed wavelength \citep{catelan04}.
 \citet{madore12} has recently shown how the intrinsic scatter of the PL relation
 decreases  with increasing
wavelength. Both reddening effects and metallicity dependencies also
decrease significantly for observations further into the IR.

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision.
\citet{madore12}
For example, using 5 stars with trigonometric parallaxes
  measured by Hubble \citep{benedict11}, \citet{madore12} measured  
  the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths --  to be of order
$\sim$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $2\%$ for RR
Lyrae stars out to $\sim$60~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve $\sim$25\% uncertainties \citep[e.g.,][]{deason12b}.

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia01}, planned for launch in late
13, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
during the 5-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances up to $\sim$6~kpc for RR Lyrae stars.

Figure~\ref{fig:gaia_errors} show the Gaia end-of-mission distance and
tangential velocity error estimates for RR Lyrae. The velocity uncertainty depends on the magnitude of the source, which depends on the metallicity: each line in Figure~\ref{fig:gaia_errors} is computed by Monte Carlo sampling from the empirical metallicity distribution of the Galactic halo from \cite{ivezic08} (centered around [Fe/H] $\sim$ -1.5). Within 2~kpc, Gaia
will be able to measure distances to these stars with better than 2\%
accuracy---RR Lyraes in this volume will be used to test and calibrate
the Spitzer PL relation described above. Beyond the 2~kpc threshold,
the mid-IR PL relation for RR Lyraes will provide more accurate
distance measurements.

@@

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<$10~kpc to 80~kpc. This enormous increase in volume will greatly refine
data on existing debris systems such as the Sagittarius stream.

@@

Most recently, \citet[][hereafter LM10]{law10} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} shows particle positions from the final time-step of an N-body simulation of dwarf satellite disruption along the expected Sgr
orbit in the best-fitting Milky Way halo model from LM10. [TODO: Maybe not true...: The positions are shown in a coordinate system defined
by the orbit of the satellite \citep{majewski04}]. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history:  
its mass and stellar populations, orbit, rate of destruction, and
even original distribution in populations [other refs??]. The phase-space distribution of the debris offers a unique probe of
the depth, shape, extent, and orientation of the Milky Way's dark
matter halo that cannot be matched using other techniques on our own
or other galaxies \citep[e.g.,][]{ibata01}. The LM10 constraints on
Sgr's mass and orbit have invigorated discussions of Sgr in a more
cosmological context, the mass and extent of the original dark matter
halo that hosted it, as well as its effect on the Milky Way itself
\citep{bailin03,purcell11,micheldansac11,gomez12}.

Combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris will open up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc ($\epsilon_{d}=0.6~{\rm kpc}$) is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\epsilon_{v}\sim2$~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim10$~km/s;][]{majewski04}, and ground-based radial
velocities, the measured phase-space
position of each individual RR Lyrae becomes a
powerful probe of the potential (see \ref{fig:}). The next section outlines a new
method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day phase-space position of the
progenitor system is also known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time
(imagine watching a visualization of a simulation of satellite destruction "rewind"), the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system (Figure~\ref{fig:ps_distance}).

This approach was originally proposed by \citep{johnston99a} and was designed and tested for the proposed characteristics of the Space Interferometry Mission \citep{unwin??}, which was expected to be able to measure proper motions with $\mu$as/year accuracy for stars as faint as $V=20$. Distances to stream members would only be poorly known and were estimated in the method by using anticipated properties of tidal debris. Below we present an updated version of the algorithm: the promise of 2\% distances to RR Lyrae stars (see Section \ref{sec:spitzer}) enables a direct measurement (rather than approximate estimate) of the position of a star within its debris structure. The test statistic that quantifies how well stars recombine with the satellite has been rigorously redefined.

@@

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,;\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the orbital radius of the satellite, $R$, and the computed enclosed mass
of the potential within $R$, $M_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=R\Big(\frac{m_{\rm sat}}{3M_{\rm enc}}\Big)^{1/3}\,\,\,\,;\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}
Qualitatively, when the distance in this normalized six-dimensional space,
$D_{{\rm ps},i}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
likely recaptured by the satellite. JZSH imposed a similar condition as a hard boundary and used the number of recaptured particles in a given backwards-integration as a cost function for optimization. What follows is a description of an updated procedure with a more natural, statistically-motivated choice for an objective function.

@@

For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars: in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way. If
\begin{equation}
  f(\bs{\theta})=\log \det(\mathrm{Cov}( \bs{A}_{ik})).
\end{equation}

[Still working on this section...need to figure out best way to express what I'm trying to say.] Though this is not explicitly a \emph{likelihood}, optimizing over this statistic is a very simple approximation to performing a maximum-likelihood optimization. Assuming the samples in $\bs{A}_{ik}$ are approximately normal-distributed, the variance is closely related to the Shannon entropy \citep{shannon1948}, $H$,  
\begin{equation}
  H \propto \log \sigma .
\end{equation}
Minimizing the entropy is then equivalent to minimizing the loss of \emph{information} for a given choice of parameters, $\bs{\theta}$, given the ``data,'' $\bs{A}_{ik}$, and is thus related to maximum-likelihood estimation \citep[see][for more detail]{mackay2002}.

>>>>

There is a long tradition for using RR Lyrae stars in the Galaxy to study structure
\citep[going back to first estimates of the distance to the Galactic center][]{shapley18},
substructure \citep[e.g.][]{sesar10}, and distances to satellite galaxies \citep[e.g.][]{clementini03}.
However, studies of RR Lyrae at optical wavelengths are limited by both metallicity affects on the intrinsic 
brightness of these stars and variable extinction along the line of sight.
Moreover, systematic differences between instruments make it difficult to tie observations across the sky
to a common scale. While RR Lyrae are among the best distance indicators available to us today, the absolute uncertainty in
individual estimates are typically $\gtrsim$10\%.

At longer wavelengths, RR Lyrae promise tighter constraints on distances.
The near-infrared period-luminosity (PL) relation for RR
Lyrae stars was first mapped by \citet{longmore86} 
using $K$-band observations of stars in Galactic
globular clusters and later shown to be due to
the
changing bolometric correction for these stars as a function of the
observed wavelength \citep{catelan04}.
 \citet{madore12} have recently shown that the intrinsic scatter of the PL relation
 decreases  with increasing
wavelength. Both reddening effects and metallicity dependencies also
decrease significantly for observations further into the IR.

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide stellar distance
indicators of unprecedented accuracy and precision.
Using five stars with trigonometric parallaxes
  measured by Hubble \citep{benedict11}, \citet{madore12} measured  
  the dispersion in the
mid-IR PL relation at Spitzer wavelengths to be $\sim$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $2\%$ for RR
Lyrae stars out to $\sim$60~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve $\sim$25\% uncertainties \citep[e.g.,][]{deason12b}.

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia01}, planned for launch in late
2013, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
during the five-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances up to $\sim$6~kpc for RR Lyrae stars.

Figure~\ref{fig:gaia_errors} shows the Gaia end-of-mission distance and
tangential velocity error estimates for RR Lyrae. The velocity uncertainty depends on the magnitude of the source, which depends on the metallicity: each line in Figure~\ref{fig:gaia_errors} is computed by Monte Carlo sampling from the empirical metallicity distribution of the Galactic halo from \cite{ivezic08} (centered around [Fe/H] $\sim$ -1.5). Within 2~kpc, Gaia
will measure distances to these stars with better than 2\%
accuracy -- RR Lyrae in this volume will be used to test and calibrate
the Spitzer PL relation described above. Beyond the 2~kpc threshold,
the mid-IR PL relation for RR Lyrae will provide more accurate
distance measurements.

@@

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs.\footnote{\url{http://www.rssd.esa.int/index.php?page=Science_Performance&project=GAIA}} The
combination of Spitzer and Gaia data will extend the ``horizon'' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<$10~kpc to 80~kpc. This enormous increase in volume will greatly refine
data on existing debris systems such as the Sagittarius stream.

@@

Most recently, \citet[][hereafter LM10]{law10} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} shows particle positions (in Galactocentric coordinates) from the final time-step of an N-body simulation of dwarf satellite disruption along the expected Sgr
orbit in the best-fitting Milky Way halo model from LM10. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history:  
its mass and stellar populations, orbit, rate of destruction, and
even original distribution in populations. The phase-space distribution of the debris also offers a unique probe of
the depth, shape, extent, and orientation of the Milky Way's dark
matter halo that cannot be matched using other techniques on our own
or other galaxies \citep[e.g.,][]{ibata01}. The LM10 constraints on
Sgr's mass and orbit have invigorated discussions of Sgr in a more
cosmological context, the mass and extent of the original dark matter
halo that hosted it, as well as its effect on the Milky Way itself
\citep{bailin03,purcell11,micheldansac11,gomez12}.

Combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris will open up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc ($\epsilon_{d}=0.6~{\rm kpc}$) is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\epsilon_{v}\sim2$~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim10$~km/s;][]{majewski04}, and ground-based radial
velocities, the measured phase-space
position of each individual RR Lyrae becomes a
powerful probe of the potential. The next section outlines a new
method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day phase-space position of the
progenitor system is also known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time
(imagine watching satellite destruction in ``rewind"), the
stars should recombine with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system (Figure~\ref{fig:ps_distance}).

This approach was originally proposed by \citep{johnston99a} and was designed and tested for the proposed characteristics of the Space Interferometry Mission \citep{unwin08}, which was expected to be able to measure proper motions with $\mu$as/year accuracy for stars as faint as $V=20$. Distances to stream members would only be poorly known and were estimated in the method by using anticipated properties of tidal debris. Below we present an updated version of the algorithm: the promise of 2\% distances to RR Lyrae stars (see Section \ref{sec:spitzer}) enables a direct measurement (rather than approximate estimate) of the position of a star within its debris structure. The test statistic that quantifies how well stars recombine with the satellite has been rigorously redefined.

@@

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,,\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the orbital radius of the satellite, $R$, and the computed enclosed mass
of the potential within $R$, $M_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=R\Big(\frac{m_{\rm sat}}{3M_{\rm enc}}\Big)^{1/3}\,\,\,\,,\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}
Qualitatively, when the distance in this normalized six-dimensional space,
$D_{{\rm ps},i}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
likely recaptured by the satellite. \citet{johnston99a} (hereafter JZSH) imposed a similar condition as a hard boundary and used the number of recaptured particles in a given backwards-integration as a cost function for optimization. What follows is a description of an updated procedure with a more natural, statistically-motivated choice for an objective function.

@@

For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of the distribution of minimum phase-space vectors
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars: in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution -- computed for a given
set of potential parameters, $\bs{\theta}$ -- is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way 
\begin{align}
  \Sigma_n &= \mathrm{Cov}( \bs{A}_{ik}) \\
  f(\bs{\theta}) &= \ln \det \Sigma.
\end{align}

[Still working on this bit...need to figure out best way to express what I'm trying to say.] Though this is not explicitly a \emph{likelihood}, optimizing over this statistic is a very simple approximation to performing a maximum-likelihood optimization. Assuming the samples in $\bs{A}_{ik}$ are approximately normal-distributed, the variance is closely related to the Shannon entropy \citep{shannon1948}, $H$,  
\begin{equation}
  H \propto \log \sigma .
\end{equation}
Minimizing the entropy is then equivalent to minimizing the loss of \emph{information} for a given choice of parameters, $\bs{\theta}$, given the ``data,'' $\bs{A}_{ik}$, and is thus related to maximum-likelihood estimation \citep[see][for more detail]{mackay2002}.
